---
title: Using the CDC COVID Socrata API in R
author: Christian Testa
date: "`r format(Sys.time(), '%b %m %Y')`"
---

The CDC Covid Socrata API is accessible here with some accompanying documentation.  
<https://dev.socrata.com/foundry/data.cdc.gov/ks3g-spdg> 

In order to use the Socrata API, one must provide their email, password, and app token for 
their account set up at <https://data.cdc.gov/login>.

To use the example R code below, specify your email, password, and app token in the 
accompanying files `data_cdc_email.txt`, `data_cdc_password.txt`, and `app_token.txt` 
on the first line of each file.

```{r setup, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  cache=TRUE
)
```


```{r load socrata}
# if RSocrata is not available, install it
if (!require(RSocrata)) install.packages("RSocrata")

# load socrata api interface
library(RSocrata)
```

#### Read in Personal/Secret Data -- Email, Password, App Token

```{r load personal/secret data}
app_token <- readLines("app_token.txt")[[1]]
data_cdc_password <- readLines("data_cdc_password.txt")[[1]]
data_cdc_email <- readLines("data_cdc_email.txt")[[1]]
```

#### Make a request through RSocrata to data.cdc.gov

```{r make request}
df <- read.socrata(
  "https://data.cdc.gov/resource/ks3g-spdg.json",
  app_token = app_token,
  email     = data_cdc_email,
  password  = data_cdc_password
)
```

#### Take a look at the data frame

```{r example features}
dim(df)
colnames(df)
unique(df$data_as_of)
unique(df$start_week)
unique(df$end_week)
unique(df$age_group)
unique(df$race_and_hispanic_origin)

str(df)
```

#### Try a Historical Request

Annoying news: If I try to make a request for the data as of a prior date, it appears that the 
request returns an empty data frame.


```{r make back-dated request}
may_1_df <- read.socrata(
  "https://data.cdc.gov/resource/ks3g-spdg.json?data_as_of=2020-05-01T00:00:00.000",
  app_token = app_token,
  email     = data_cdc_email,
  password  = data_cdc_password,
)

may_1_df
```

### Plotting Total Deaths by Race and Hispanic Origin

I want to be extra clear here: depicted below are deaths from all causes of death, not exclusively 
COVID-19 related. This is just for purposes of illustrating some of the information the data contains. 

```{r, fig.height = 7, fig.width = 14}
library(dplyr)
library(ggplot2)

all_ages_us_df <- df %>% 
  filter(age_group == 'All Ages',
         state == 'United States') %>% 
  mutate(total_deaths = as.numeric(total_deaths)) 

ggplot(all_ages_us_df, aes(x = race_and_hispanic_origin, y = total_deaths, fill = race_and_hispanic_origin)) + 
  geom_bar(stat='identity') + 
  expand_limits(y = 0) + 
  coord_flip() + 
  theme_minimal() + 
  scale_y_continuous(name="Total Deaths", labels = scales::comma) + 
  xlab("") + 
  ggtitle("Total Deaths by Race/Ethnicity as of May 20th, 2020") + 
  guides(fill=guide_legend(title = "Race and Hispanic Origin", nrow=4, byrow=TRUE)) + 
  theme(
    axis.text.x = element_text(angle = 45),
    legend.position = 'bottom',
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) 
```


And on log-scale:

```{r, fig.height = 7, fig.width = 14}
df %>% 
  filter(age_group == 'All Ages',
         state == 'United States') %>% 
  mutate(total_deaths = as.numeric(total_deaths)) %>% 
ggplot(aes(x = race_and_hispanic_origin, y = total_deaths, fill = race_and_hispanic_origin)) + 
  geom_bar(stat='identity') + 
  expand_limits(y = 0) + 
  coord_flip() + 
  theme_minimal() + 
  scale_y_log10(name="Total Deaths", labels = scales::comma) + 
  xlab("") + 
  ggtitle("Total Deaths by Race/Ethnicity as of May 20th, 2020 - Log Scale") + 
  guides(fill=guide_legend(title = "Race and Hispanic Origin", nrow=4, byrow=TRUE)) + 
  theme(
    axis.text.x = element_text(angle = 45),
    legend.position = 'bottom',
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank()
  ) 
```
